Efficient Computation of Extended Sources

### 2020-11-23 repro/papers/sseprecond/paper.tex
more

The class of non-point sources investigated here are those confined a hypersurface assumed to contain the locations of physical (point) sources, but not a priori required to have point support. These {\em surface extended sources} are able to yield a relatively small residual in the source subproblem \ref{eqn:esis} for more or less arbitrary coefficient array $c$. This observation is a by-product of the approximate solution for \ref{eqn:esis} constructed below. The construction of the approximate solution is closely related to the {\em time reversal} method of photoacoustic tomography (see \cite{StefanovUhlmannIP:09} and references cited there).

The next section gives a precise description of an acoustic version of problem \ref{eqn:esis}, and points out two interesting special cases, the {\em crosswell} and {\em diving wave} configurations. The assumptions detailed here imply that the data $d$ of the inverse problems \ref{eqn:esi}, \ref{eqn:esis} have the physical character of {\em transmitted waves}. There are no mathematical results at present concerning reflected wave inverse problems treated by source extension methods, though there are tantalizing numerical clues \cite[]{LeeuwenHerrmannWRI:13,Warner:14,Warner:16,LeeuwenHerrmann:16,HuangSymes:Geo17,HuangSymes:Geo18a,HuangSymes:Geo18b}.

@@

%The penalty operator $A$ is chosen to penalize non-physical aspects of the trial source $g$: physical sources are supposed to constitute the null space of $A$.

%In some cases, the resulting  relaxed data fit problem has been observed to be more amenable to solution by local optimization. The main goal of this paper is explain why this should be so,  in a particular instance of source-extended FWI. I will show that for a specific choice of {\em annihilator} $A$, the optimization problem \ref{eqn:esi} possesses a {\em qualitatively} larger domain of convexity for consistent data ($PL[c]^{-1}g = d, Ag=0$) than does the FWI problem \ref{eqn:ols}, while possessing the same solution.

%For sake of argument (and because almost all work so far has assumed it), adopt the isotropic point source as the model: for the data gather at source positon $\bx_s$, the source takes the form
%\begin{equation}
%\label{eqn:ptsrc}
%f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s). 
%\end{equation}
%in which the {\em wavelet} $w(t)$ has whatever structure is necessary for $f$ to lie in the range of the wave operator $L[c]$. FWI can also accommodate more complex source models reflecting the nontrivial radiation patterns of actual sources. 
%The data projection operator $P$ involves the restriction of one or more components of the field $u$ to a hypersurface $X_r$. The simplest interesting choice of $X_r$ is a hyperplane, say $z=z_r$ (with the convention that $\bx=(x,y,z)$). To emulate the limited aperture of any experimental data system, include multiplication by a smooth cut-off or taper function $\phi$. Component selection is a special case of multiplication by a matrix $P_c \in \bR^{n_d \times n_s}$, in which $n_d$ is the number of data components and $n_s$ is the dimension of the state vector $u$.
%\begin{equation}
%\label{eqn:data}
%Pu(x,y,t) = \phi(x,y,z_r)P_cu(x,y,z_r,t)
%\end{equation}
%in which $P_c$ is a

%Source extension replaces $f(\bx,t;\bx_s)$ with an artificial source $g(\bx,t; \bx_s)$ with more degrees of freedom than merely a single common source wavelet $w(t)$ as in definition \ref{eqn:ptsrc}. Assuming that \ref{eqn:ptsrc} is actually correct physics, these extra degrees of freedom have to be suppressed in the optimal solution, via an {\em annihilator} $A$, an operator acting on extended sources $g$, whose null space is precisely the physical sources (of the form \ref{eqn:ptsrc}). Extended source inversion replaces the FWI objective with a penalty function

%The way in which degrees of freedom are introduced, and the choice of annihilator, distinguish the various source extension methods, of which there are many - see \cite{HuangNammourSymesDollizal:SEG19}, who introduce the {\em surface source} extension. It is this surface source extension that is the topic of this paper. It is similar  to the volume extension \cite[]{HuangSymes:Geo18b}, but the source is spread over the source space-time surface, rather than over space at $t=0$.
%Similar to the space-time and volume extensions \cite[]{HuangSymes:Geo18a,HuangSymes:Geo18b}, the annihilator is given by the {\em localization penalty}:
%\begin{equation}
%\label{eqn:msste}
%A[g](\bx,t;\bx_s) = |\bx-\bx_s|g(\bx,t;\bx_s). 
%\end{equation} 
%Therefore an estimate of the the source wavelet is a by-product of the inversion, and does not need to be specified {\em a priori}. Like the volume extension (and unlike, for example, Wavefield Reconstruction Inversion \cite[]{LeeuwenHerrmannWRI:13,LeeuwenHerrmann:16}), the additional data volume required is on the same order of size as the data.

%The volume and surface extensions have a well-defined relation to tomographic principles (slope or traveltime, depending on type of annihilator). This relation holds regardless of ray geometry (with or without triplication) in transmission configuration.

%Again, nothing is known theoretically about the behaviour of any of these methods for reflection data, and numerical evidence is ambiguous.

\section{Extended Source Modeling}
For acoustic wave physics, the coefficient vector $c$ is the pair $c=(\kappa,\beta)$ of bulk modulus $\kappa$ and buoyancy (reciprocal density) $\beta$, and the state vector $u=(p,\bv)$ consists of pressure $p$ (a scalar space-time field) and particle velocity $\bv$ (a vector space-time field). The wave operator $L[c]$ is given by the linear acoustics system:
\begin{equation}
\label{eqn:awe}
L[c]u = 
\left(
\begin{array}{c}
\partial_t p  + \kappa \nabla \cdot \bv \\
\partial_t \bv + \beta \nabla p
\end{array}
\right) 
\end{equation}

The space dimension will be denoted by $d$. Most of what follows is valid for any $d >0$. The choice $d=2$ is computationally convenient, if non-physical.

@@

Real-world sampling is discrete and direction-sensitive. This paper ignores such issues and presumes that the sampling operator $P$ simply restricts its argument to a cylindrical space-time hypersurface $\Gamma_r \times \bR$.  Here $\Gamma_r \subset \bR^d$ is an open subset of a smooth codimension 1 oriented hypersurface. To distinguish this output or receiver sampling operator from another similar operator, add the subscript $r$ to the notation, that is, write $P_r$ rather than $P$. For technical reasons (and to emulate actual acquisition practice) include multiplication by a cutoff or mute $\phi_r \in C_0^{\infty}(\bR^{d+1})$, and extract only samples of the pressure field:
\begin{equation}
\label{eqn:recp}
(P_ru)(\bx,t) = \phi_r(\bx,t)p(\bx,t),\,\bx \in \Gamma_r, t \in \bR. 
\end{equation}
A similar sampling operator for the velocity field will also be required below. Denote by ${\bf n}_r$ the outward unit normal field on $\Gamma_r$, and define
\begin{equation}
\label{eqn:recv}
(V_ru)(\bx,t) = \phi_r(\bx,t){\bf n}_r(\bx)\cdot \bv(\bx,t),\,\bx \in \Gamma_r, t \in \bR.
\end{equation}

Physical sources are modeled as an isotropic constitutive law defects of point support. As mentioned earlier, it is simple to accommodate more complex source models. Assume that all source lie in an open subset $\Gamma_s \subset \bR^d$ of a smooth orientable codimension 1 hypersurface in $\bR^d$, possibly different from the hypersurface containing $\Gamma_r$. Denote by $P_s, V_s$ sampling operators on $\Gamma_s \times \bR$ defined precisely analogously to $P_r$ and $V_r$ (equations \ref{eqn:recp}, \ref{eqn:recv}):
\begin{equation}
\label{eqn:recp}
(P_su)(\bx,t) = \phi_s(\bx,t)p(\bx,t),\,\bx \in \Gamma_s, t \in \bR. 
\end{equation}
and similarly for $V_s$. Denote by $\delta_{\bx_s}$ the delta distribution on $\Gamma_s$ at $\bx_s \in \Gamma_s$, and by $w \in C_0^{\infty}(\bR)$ a choice of source wavelet. The corresponding physical source is then the right-hand side in the acoustic system
\begin{equation}
\label{eqn:awe-src}
L[c]u = P_s^T(w\delta_{\bx_s}) 
\end{equation}
With dummy space and time variables, the right hand side above is exactly the form $w(t)e_1\delta(\bx-\bx_s)$ given in the point source definition \ref{eqn:ptsrc}, where , ${\bf e}_1$ the standard basis vector $e_1=(1,0,...,0) \in \bR^{d+1}$.

As explained in the introduction, unless the data $d$ and the coefficient array $c$ are compatible, the source subproblem \ref{eqn:esis} has no low-residual solution of the form $f = P_s^T(w\delta_{\bx_s}) $. A larger class of source models is required to fit data. A precise description of the class described in the introduction is simply the range of $P_s^T$: that is, for $h \in C^{\infty}(\Gamma_s \times \bR)$, define
\begin{equation}
S[c]h = P_rL[c]^{-1}P_s^T h.
\end{equation}
$S[c]$ is the surface source extension modeling operator. For this class of sources, the source subproblem becomes 
\begin{equation}
\label{eqn:esis-sse}
\mbox{given } c, \mbox{ choose } h  \mbox{ to minimize } \|S[c]h -d \|^2 + \alpha \|Ah\|^2 
\end{equation}
To complete the definition of this problem, a choice of {\em annihilator} $A$ is required. A simple choice is
\begin{equation}
\label{eqn:msste}
Ah(\bx,t) = |\bx-\bx_s|h(\bx,t). 
\end{equation}

Formally, the solution of the source subproblem is also the solution of the normal equation
\begin{equation}
\label{eqn:normal}
(S[c]^TS[c] + \alpha^2 A^TA) h = S[c]^T d 
\end{equation}
This relation holds, and a solution $h$ is guaranteed to exist, if the $\alpha=0$ normal operator $S[c]^TS[c]$ is well-defined and continuous in $L^2(\bar{\Gamma}_s \times \bR)$, and there exists $C>0$ so that 
\begin{equation}
\label{eqn:coercive}
\|S[c]h\|^2 \le C|h|^2,\,\,\|S[c]h\|^2 + \alpha^2 \|Ah\|^2 \ge C^{-1}\|h\|^2
\end{equation}
for all $h \in L^2(\bar{\Gamma}_s)$. These estimates hold under various conditions. One particular setting in which the least squares problem \ref{eqn:esis} is well-posed is given by

\noindent {\bf Assumption A:} There exists $\theta_*>0$ so that for every ray connecting a point in $\mbox{supp }\phi_s$ with a point in $\mbox{supp }\phi_r$, the angles $\theta_s$ and $\theta_r$ subtended by the ray and the tangent planes of $\Gamma_s$ and $\Gamma_r$ respectively are both greater than $\theta_*$ in absolute value.

The first appendix provides a proof of the following statement:

\begin{theorem} 
\label{thm:one}
Suppose that $\log \kappa, \log \beta \in C^{\infty}(\bR^d) \cap L^{\infty}(\bR^d)$, and that Assumption A holds. Then the estimates \ref{eqn:coercive} hold for $h \in C^{\infty}_0(\Gamma_s \times \bR)$, hence $S[c]$ extends to $L^2(\bar{\Gamma}_s \times \bR)$ with the same estimates.
\end{theorem}

The assumptions made in the hypotheses of \ref{thm:one} imply that the inverse wave problems under consideration in this paper are of the transmission type, that is, do not involve reflected waves. The reflected wave concept is well-defined only in terms of high-frequency asymptotics: because the coefficients are assumed smooth, there are no material singularities to generate reflections.  Two configurations of sources and receivers that naturally fit this framework, at least in an ideal sense, are
\begin{itemize}
\item the {\em crosswell} configuration, in which $\Gamma_s$ and $\Gamma_r$ are parallel hyperplanes. In reality, wells are approximately 1D structures embedded in 3D ambient media, so this configuration does not really model them well. However it has become commonplace to use the 2D version of this configuration as a model, assuming that the waves propagate in-plane between the wells  \cite[]{BregChapBai:86,Harr:93,Zhouetal:93,Pratt:99b,Plessix:00}.
\item the {\em diving wave} configuration, $\Gamma_s$ and $\Gamma_r$ are both subsets of a horizontal plane at the source, respectively receiver, depth (assumed well-defined). In marine towed streamer and ocean bottom seismometer surveys, a set of refracted waves arrives earlier than the direct wave in the water column, at sufficient distance from the source. The mutes $\phi_s, \phi_r$ isolate these refacted (``diving wave'') arrivals. This data is the focus of the ``classic'' period of FWI \cite[]{SirgPratt02,BrendersPratt:06a,BrendersPratt:06b,VirieuxOperto:09,Vigh:10,Vigh13}.
\end{itemize}
In both cases, it is possible (though not necessary) that Assumption A is satisfied. Note that in the diving wave configuration, the mutes $\phi_S,\phi_r$ are responsible for eliminating direct wave energy, hence enforcing the acute incidence of the waves connecting the admitted source and receiver points. The numerical examples presented below will fall into the crosswell category.

Assumption A is not by any means the most general hypothesis under which the source subproblem has a well-behaved solution. However weaker hypotheses typically involve multidimensional filters applied to data (and extended sources) to eliminate waves propagating tangent to the source and receiver surfaces, so require more involved implementations.

%Denote by $J_{\alpha}[g,c;d]$ the objective function defined in \ref{eqn:esi}:
%\begin{equation}
%\label{eqn:jdef}
%J_{\alpha}[g,c;d] = \frac{1}{2}( \|PL[c]^{-1}g -d \|^2 + \alpha \|A[g]\|^2) 
%\end{equation}
%To compute the objective 
%\begin{itemize}
%\item solve the acoustic wave system \ref{eqn:awe} with $g$ as in \ref{eqn:mssur}
%\item sample at receiver locations and compute mean-square difference with pressure data traces, take mean square
%\item add scaled mean square of annihilator applied to the source
%\end{itemize}

%The variable projection method \cite[]{GolubPereyra:03} applied to the objective \ref{eqn:esi} approaches its minimization via a nested algorithm: first minimize \ref{eqn:esi} over the source $g$ for given coefficients $c$, thus making $g = g[c]$ a function of $c$ (inner minimization); then substitute $g[c]$ in $J$ and minimize the resulting function of $c$ (outer minimization). The optimal $g[c]$ solves the first order necessary condition for minimization of $J$ over $g$: $\nabla_g J_{\alpha}[g,c;d] = 0$. For linear annihilators such as that defined by \ref{eqn:mssur}, $J$ is quadratic in $g$, so the first order necessary condition is also sufficient, and is equivalent to the linear {\em normal equation}:

%where I have written $S[c]=PL[c]^{-1}$ for the operator that maps the (extended) source to the data traces. Substitution of $g[c]$ in $J$ results in the {\em reduced objective}
%\begin{equation}
%\label{eqn:jred}
%\tilde{J}_{\alpha}[c;d] = J_{\alpha}[g[c],c;d].
%\end{equation}
%The gradient would appear to involve the derivative of the inner solution $g[c]$ with respect to $c$, computation of which implies solution of additional linear systems. However by fortunate accident, the chain rule and the normal equation \ref{eqn:normal}
%conspire to eliminate all terms involving derivatives of $g[c]$. Consequently the direction derivative of $\tilde{J}_{\alpha}$ in the direction $\delta c$ is
%\begin{equation}
%\label{eqn:tildejderiv}
%D_c\tilde{J}_{\alpha}[c;d]\delta c = \langle D(S[c]g)\delta c,S[c]g-d \rangle|_{g=g[c]}
%\end{equation}
%whence
%\begin{equation}
%\label{eqn:l2grad}
%\nabla \tilde{J}_{\alpha}[c;d] = D_c(S[c]g)^T|_{g=g[c]}(S[c]g[c]-d)
%\end{equation}
%Let us field-strip the expression on the right. The rightmost factor is the data residual, which is a by-product of many methods for the solution of the normal equation \ref{eqn:normal}. $D_cS$ is the linearization or Born approximation of $S$, or perturbation with respect to $c$, with $g$ fixed:
%\[
%S[c+\delta c]g \approx S[c]g + D_c(S[c]g)\delta c
%\]
%That is, $D_c(S[c]g)$ is the object commonly called the Born modeling operator, with source wavefield $g$. Its adjoint or transpose is $D_c(S[c]g)^T$ - as is well known, this is the RTM operator (or rather a version of it), computed via the adjoint state method in the time domain. Note that the Born and RTM operators need to match, i.e. both be derived from the same underlying full waveform modeling operator (in this case acoustodynamics, but the same would apply for any other wave theory).

%The upshot is that the computation of the gradient goes like this:
%\begin{itemize}
%\item solve the normal equation \ref{eqn:normal} for $g[c]$
%\item compute (or retain) the data residual $S[c]g[c]-d$
%\item reverse-time migrate the data residual, using $g[c]$ to generate the source wavefield in the adjoint state method - the output is the gradient of $\tilde{J}_{\alpha}$.
%\end{itemize}

%To make this algorithm practical and efficient, effective preconditioning needs to be applied to both inner and outer problems.

\subsection{Time Reversal}
%In contrast to the outer problem, a preconditioner for the inner problem is available that works regardless of medium smoothness, so for both transmission and reflection (the latter with some caveats). 
%First the transmission case - a sort of rotated crosswell configuration. Idealize the receiver set as a continuum plane $z=z_r$, with $z_r > z_s$ - not essential but simpler.

%NB: the definition of the forward map (and the data) should include a taper $\phi(\bx_r,\bx_s)$ masking out the shot gather $d_r(\cdot,\cdot;\bx_s)$ for each source position $\bx_s$. However I will not explicitly mention this operation.

Change notation slightly, and write $S[c]h_s$ rather than $S[c]g$ with $g = h_s \delta(z-z_s)$, as $z_s$ will be presumed ot be the same for all shots. The pressure and velocity fields occuring in the computation of $d_r=S[c]h_s$ form the solution of
\begin{eqnarray}
\label{eqn:awedata}
\frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv +
h \delta(z-z_s), \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t<0\nonumber\\ 
\bv & = & 0 \mbox{ for } t<0 \nonumber\\
 p|_{z=z_r}& = & d_r
\end{eqnarray}
Suppose that the recording time interval $[0,T]$ is long enough that $ d_r \approx 0$ for $t > T$. Note that $p,\bv$ are {\em incoming} (downgoing) in $z>z_s$. Therefore in $z<z_r$ {\em the same fields} approximately solve
\begin{eqnarray}
\label{eqn:awetr}
\frac{\partial p}{\partial t} & = & - \kappa \nabla \cdot \bv \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t>T\nonumber\\ 
\bv & = & 0 \mbox{ for } t>T \nonumber\\
p|_{z=z_r} &=&  d_r
\end{eqnarray} 
The last equation defines $p$ on the entire plane $z=z_r$ for $t<T$, and the previous two define final conditions at $t=T$. So the {\em time-reversed} system \ref{eqn:awetr} has unique solutions in $z<z_r, t<T$ and $z>z_r,t<T$. The solution in $z<z_r$ is incoming (downgoing) at $z=z_s$, and is {\em the same vector of fields} (approximately) as the solution of \ref{eqn:awedata} in the region $z_s<z<z_r$ between the source and receiver surfaces.

>>>>

The class of non-point sources investigated here are those confined a hypersurface assumed to contain the locations of physical (point) sources, but not {\em a priori} required to have point support. These {\em surface extended sources} are able to yield a relatively small residual in the source subproblem \ref{eqn:esis} for more or less arbitrary coefficient array $c$. This observation is a by-product of the approximate solution for \ref{eqn:esis} constructed below. The construction of the approximate solution is closely related to the {\em time reversal} method of photoacoustic tomography (see \cite{StefanovUhlmannIP:09} and references cited there).

The next section gives a precise description of an acoustic version of problem \ref{eqn:esis} for an idealized {\em crosswell} configuration, in which source located on a surface $z=z_s$ generate waves that propagate to to a receiver surface $z=z_r > z_s$. In principle, {\em diving waves} can be described with similar mathematics, but with some additional complications, so I do not treat the diving wave case explicitly in this paper. The assumptions detailed here imply that the data $d$ of the inverse problems \ref{eqn:esi}, \ref{eqn:esis} have the physical character of {\em transmitted waves}. There are no mathematical results at present concerning reflected wave inverse problems treated by source extension methods, though there are tantalizing numerical clues \cite[]{LeeuwenHerrmannWRI:13,Warner:14,Warner:16,LeeuwenHerrmann:16,HuangSymes:Geo17,HuangSymes:Geo18a,HuangSymes:Geo18b}.

@@

\section{Operators and Adjoints}

For acoustic wave physics, the coefficient vector $c$ is the pair $c=(\kappa,\beta)$ of bulk modulus $\kappa$ and density $\rho$, and the state vector $u=(p,\bv)^T$ consists of pressure $p$ (a scalar space-time field) and particle velocity $\bv$ (a vector space-time field). The wave operator $L[c]$ is given by the linear acoustics system:
\begin{equation}
\label{eqn:aweop}
L[c]u = 
\left(
\begin{array}{c}
\frac{1}{\kappa}\frac{\partial p}{\partial t}  + \nabla \cdot \bv, \\
\rho\frac{\partial \bv}{\partial t} + \nabla p.
\end{array}
\right) 
\end{equation}
That is,
\begin{equation}
  \label{eqn:awemat}
  L[c] = \left(
    \begin{array}{cc}
      \frac{1}{\kappa}\frac{\partial}{\partial t} & \nabla \cdot \\
      \nabla & \rho \frac{\partial}{\partial t}
    \end{array}
  \right)
\end{equation}

The space dimension is $d$. Most of what follows is valid for any $d >0$. Examples later in this paper will use $d=2$ for computationally convenience.

@@

The idealized ``crosswell'' configuration involves sources confined to $z=z_s$. I will mostly represent sources as constitutive law defects, that is, right-hand sides in the system $L[c]u=f$ of the form $f(\bx,t) = (h_s(x,y,t)\delta(z-z_s), {\bf 0})^T$. With the choice $L[c]$ given in \ref{eqn:aweop}, the causal wave system $L[c]u^+=f$ takes the form
\begin{eqnarray}
\label{eqn:awep}
\frac{1}{\kappa}\frac{\partial p^{+}}{\partial t} & = & - \nabla \cdot \bv^{+} +
h_s \delta(z-z_s), \nonumber \\
\rho\frac{\partial \bv^{+}}{\partial t} & = & - \nabla p^{+},\nonumber \\
p^{+} & =& 0 \mbox{ for } t \ll 0,\nonumber\\ 
\bv^{+} & = & 0 \mbox{ for } t \ll 0.
\end{eqnarray}

Similarly, the anti-causal wave system $L[c]u^-=f$ takes the form
\begin{eqnarray}
 \label{eqn:awem}
\frac{1}{\kappa}\frac{\partial p^{-}}{\partial t} & = & - \nabla \cdot \bv^{-} +
h_s \delta(z-z_s), \nonumber \\
\rho \frac{\partial \bv^{-}}{\partial t} & = & - \nabla p^{-},\nonumber \\
p^{-} & =& 0 \mbox{ for } t \gg 0,\nonumber\\ 
\bv^{-} & = & 0 \mbox{ for } t \gg 0.
\end{eqnarray}
Sampling the causal pressure field $p^{+}$ on the surface $z=z_r$ defines the causal forward modeling operator $S^{+}_{z_s,z_r}[c]$:
\begin{equation}
  \label{eqn:fwd}
  S^{+}_{z_s,z_r}[c]h_s = p^+|_{z=z_r}
\end{equation}
The anti-causal forward modeling operator $S^-_{z_s,z_r}$ has a similar definition.
The subscript signifies that sources are located on $z=z_s$, the receivers on $z=z_r$. It's necessary to include this information in the notation, as versions of $S$ with sources and receivers in several locations will be needed in the discussion below.

The adjoint of $S^+_{z_s,z_r}[c]$ can be computed
by a variant of the adjoint state method, in this case a by-product of the conservation of energy. Suppose that $p^-,\bv^-$ solve \ref{eqn:awem} with $h_s\delta(z-z_s)$ replaced by $ h_r \delta(z-z_r)$. Then
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  
\rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow \infty}
-
\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)|_{t \rightarrow -\infty}
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p^+ p^-}{\kappa} +  \rho \bv^+ \cdot \bv^- \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p^+}{\partial t} p^- +  p^+ \frac{1}{\kappa}\frac{\partial p^-}{\partial t} \right.
\]
\[
+
\left. \rho \frac{\partial \bv^+}{\partial t} \cdot \bv^- + \rho \bv^+ \cdot \frac{\partial \bv^-}{\partial t} \right)
\]
\[
= 
\int_{-\infty}^{\infty} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  (- \nabla p^+) \cdot \bv^- + \bv^+ \cdot (-\nabla p^-) \right)
\]
\[
= 
\int_{-\infty}^{\infty}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv^+ + 
 h_s \delta(z-z_s)\right) p^- + p^+ \left(- \nabla \cdot \bv^- + 
 h_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  p^+ (\nabla \cdot \bv^-) + (\nabla \cdot \bv^+) p^- \right)
\]
after integration by parts in the last two terms. Most of what is left cancels, leaving 
\[
0 = \int_{-\infty}^{\infty}\,dt\,dx\,dy\, h_sp^-|_{z=z_s} +  h_rp^+|_{z=z_r} = \langle h_s, p^-|_{z=z_s}\rangle + \langle h_r, S[c]^+_{z_s,z_r}[c]h_s \rangle
\]
whence
\begin{equation}
\label{eqn:sadj}
 S^+_{z_s,z_r}[c]^T h_r = -p^-|_{z=z_s}
\end{equation}

The systems \ref{eqn:awep} and \ref{eqn:awem} differ only in the direction of time evolution (causal vs. anti-causal). Define $R$ to be the {\em time-reversal operator} on functions of space-time,
$Rf(\bx,t) = f(\bx,-t)$, and ${\cal R}$ to be the {\em acoustic field time-reversal operator} 
\begin{equation}
  \label{eqn:trdef}
  {\cal R} \left(
    \begin{array}{c}
      p\\
      \bv
    \end{array}
  \right) =
  \left(
    \begin{array}{c}
      Rp\\
      -R\bv
    \end{array}
  \right)
\end{equation}
Then ${\cal R}(p^-,\bv^-)$ solves \ref{eqn:awep} with $h_s\delta(z-z_s)$ replaced by $-Rh_r\delta(z-z_r)$. That is,
\begin{equation}
  \label{eqn:trsadj}
  Rp^-|_{z=z_s} = -S^+_{z_r,z_s}[c]Rh_r
\end{equation}
Since $R^2 = I$ and ${\cal R}^2 = I$, the identities \ref{eqn:sadj} and \ref{eqn:trsadj} imply that
\begin{equation} 
  \label{eqn:trtr}
  S^+_{z_s,z_r}[c]^T = RS^+_{z_r,z_s}[c]R.
\end{equation}

\section{Surface source-to-pressure operator}

A particular case of the relation \ref{eqn:trtr} is important in its own right, when $z_r=z_s$. Define surface source-to-pressure operator $\Lambda_{z_s}$ by
\begin{equation}
  \label{eqn:ntoddef}
  \Lambda_{z_s} = S^+_{z_s,z_s}.
\end{equation}
Then \ref{eqn:trtr} reads
\begin{equation}
  \label{eqn:ntodtr}
  \Lambda_{z_s}^T = R\Lambda_{z_s}R.
\end{equation}
This operator is closely related to the operator mapping $\lim_{z\rightarrow z_s^+} \bv_z^+$ to $\lim_{z\rightarrow z_s^+}  p^+$, often called the {\em Neumann-to-Dirichlet} operator. The connection is explained below.

The quadratic form defined by $\Lambda_{z_s}$ has fundamental physical significance. Suppose that the surface source $h_s$ vanishes outside of a bounded time interval $[t_{\rm min},t_{\rm max}]$. Then the causal acoustic field $(p^+,\bv^+)$, solution of the system \ref{eqn:awep}, vanishes for $t<t_{\rm min}$. Define the total acoustic energy $E(t)$ at time $t$ by
\begin{equation}
  \label{eqn:defae}
  E(t) = \frac{1}{2} \int \,d\bx \, \left(\frac{(p^+)^2}{\kappa} + \rho |\bv^+|^2\right)
\end{equation}
Then since $E(t_{\rm min})=0$,
\[
  E(t_{\rm max}) = \int_{t_{\rm min}}^{t_{\rm max}} \frac{dE}{dt}
\]
\[
  = \int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\, \left(p^+ \frac{1}{\kappa}\frac{\partial p^+}{\partial t} + \bv^+ \cdot \rho \frac{\partial \bv^+}{\partial t}\right)
\]
\[
  =\int_{t_{\rm min}}^{t_{\rm max}} \int\,d\bx\,\left(p^+(-\nabla \cdot \bv^{+} + h_s\delta(z-z_s)) + \bv^+ \cdot \rho \frac{\partial \bv^+}{\partial t}\right)
\]
\[
  \int_{t_{\rm min}}^{t_{\rm max}}\int\,d\bx\, \left(\left(\nabla p^+\rho \frac{\partial \bv^+}{\partial t}\right) + p^+h_s\delta(z-z_s)\right)
\]
\[
  =  \int_{t_{\rm min}}^{t_{\rm max}}\int\,dxdy h_s p^+|_{z=z_s} = \langle h_s, \Lambda_{z_s} h_s \rangle.
\]
Since the source $h_s\delta(z-z_s)$ no longer acts on the fluid for $t >t_{\rm max}$, $E_{\rm tot}= E(t_{\rm max})$ is the total energy transferred from the source to the fluid, and
\begin{equation}
  \label{eqn:surfenergy}
  E_{\rm tot} = \langle h_s, \Lambda_{z_s} h_s \rangle.
\end{equation}
We can conclude from identity \ref{eqn:surfenergy} that the symmetric part of $\Lambda_{z_s}$ is positive semidefinite:
\begin{equation}
  \label{eqn:lamsd}
  \frac{1}{2}(\Lambda_{z_s}^T + \Lambda_{z_s}) \ge 0.
\end{equation}

Observe that the causal acoustic system \ref{eqn:awep} implies directly that
\begin{equation}
  \label{eqn:sourcevjump}
  h_s= [v_z]|_{z_s}.
\end{equation}

\section{Progressing Waves and Symbols}

The symbol $\sigma(L[c])$ of a matrix operator $L[c]$ is a complex scalar-valued matrix of the same size. In terms of its product with an arbitrary vector $\bu \in \bR^{d+1}$, it is
\[
  \sigma(L[c]) \bu = e^{-i(\omega t + \bk \cdot \bx)} (L[c]e^{i(\omega t + \bk \cdot \bx)}\bu).
\]
For $L[c]$ given by the definition \ref{eqn:awemat}, obtain
\begin{equation}
  \sigma(L[c])(t,\bx,\omega,\bk) = i\left(
    \begin{array}{cc}
      \frac{\omega}{\kappa(\bx)} & \bk^T \\
      \bk & \rho(\bx) \omega)
    \end{array}
  \right)
\end{equation}

The characteristic equation of $L[c]$ is $\det \sigma(L[c]) = 0$. The significance of the characteristic equation is its role in constraining the phase $\psi(\bx)$ in the progressing wave ansatz,
\begin{equation}
  \label{eqn:go}
  \bu(\bx,t) = e^{i\omega (t-\psi(\bx))} \ba (\bx).
\end{equation}
Applying $L[c]$ to this ansatz, obtain
\[
  L[c]\bu(\bx,t)=i \omega
  \left(
    \begin{array}{c}
      \frac{1}{\kappa}a_p -  \nabla \psi \cdot \ba_v \\
      -a_p \nabla \psi + \rho \ba_v
    \end{array}
  \right) + ...
\]
\[
  = i \omega \sigma(L[c])(t,\bx,1,- \nabla \psi(\bx))ba + ...
\]
where ``$...$'' denotes terms of lower order in $\omega$. Thus $\bu$ as defined in equation \ref{eqn:go} is a high-frequency asymptotic solution, $L[c]\bu \approx 0$, if
\begin{itemize}
\item
  \begin{equation}
    \label{eqn:eik}
    \det(\sigma(L[c])(t,\bx,1,-\nabla \psi(\bx))) = 0,
  \end{equation}
  (eikonal equation), and
\item
  $\ba$ is a null vector of $\sigma(L[c])(t,\bx,1,-\nabla \psi)$.
\end{itemize}
Explicitly, the eikonal equation reads
\[
  0 = \rho^2\left(\frac{\rho}{\kappa}-|\nabla \psi|^2\right),
\]
equivalent to the familar form
\begin{equation}
  \label{eqn:usualeik}
  c|\nabla \psi| = 1,
\end{equation}
in which $c = \rho/\kappa$ is the wave velocity. Solution via the method of characteristics gives
\[
  \psi(\bx(t)) = \psi(\bx(0)) + t
\]
for a ray $t \mapsto \bx(t)$ satisfying
\[
  \frac{d\bx}{dt}(t) = -c(\bx)^2\nabla \psi(\bx(t))
\]
Finally, $\ba$ is a null vector of $\sigma(L[c])(\cdot,\cdot,1,-\nabla \psi)$ iff
\begin{equation}
  \label{eqn:godton}
  \ba_v = \frac{a_p}{\rho}\nabla \psi
\end{equation}
The scalar $a_p$ evolves along the ray according to the transport equation, which I will not derive here; suffice it to say that $a_p(t) = a_p(0) \gamma(t)$, where $\gamma > 0$ is smooth and $\gamma(0)=1$.

Suppose that $\bx_0 = (x,y,z_s)^T$ and $(\xi_x,\xi_y) \in \bR^2$ satisfies $c(\bx_0)|(\xi_x,\xi_y)|<1$. Then there is a neighborhood $\Omega$ of $\bx_0$ and $\xi_z \in C^{\infty}(\Omega), \xi_z>0$, so that $\bxi^{\pm}(\bx)=(\xi_x,\xi_y,\pm \xi_z(\bx))$ satisfy $c(\bx)|\bxi^{\pm}(\bx)|=1$ for $\bx \in \Omega$.
